---
title: "Telomere Content Analysis in the 1000 Genomes Project"
author: "Priyanshi Shah"
output: html_document
---

## Overview: This analysis investigates variation in telomere content across global populations using whole-genome sequencing data from the 1000 Genomes Project. Telomere content was estimated using qmotif-derived metrics and analyzed across populations, superpopulations, and sex.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 300)

library(tidyverse)
library(colorspace)
library(ggplot2)
library(dplyr)
library(FSA)
library(car)
library(conover.test)
library(rstatix)
```

## Global Plot Theme
```{r}

my_theme <- theme_minimal(base_size = 14) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text = element_text(face = "bold")
  )

```

---
## Data Loading & Cleaning

```{r}

telo_data <- read.csv("data/processed/1k_genome_project_analysis_csv.csv") %>%
  filter(!is.na(ScaledTelomericReads)) %>%
  mutate(
    ScaledTelomericReads = as.numeric(ScaledTelomericReads),
    Superpopulation = trimws(Superpopulation.name),
    Population = factor(Population),
    Gender = factor(Gender)
  ) %>%
  filter(ScaledTelomericReads > 0) %>%   # remove rows with scaled telomeric reads equal to 0
  mutate(LogTelomere = log(ScaledTelomericReads))


# Sanity check
unique(telo_data$Superpopulation)
```

## Color Palette

```{r}

my_palette <- qualitative_hcl(length(unique(telo_data$Population)), palette = "Dark 3")

```
---

## Visualization

### Superpopulation Distribution

```{r}
superpop_plot <- ggplot(telo_data, aes(Superpopulation, ScaledTelomericReads, fill = Superpopulation)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", color = "red", size = 3) +
  guides(fill = "none") +
  labs(x = "Superpopulations", y = "Scaled Telomeric Reads") +
  my_theme

ggsave("superpop_boxplot.png", superpop_plot, width = 10, height = 8, dpi = 300)
superpop_plot
```

---

### Population-Level Distribution

```{r}
allpop_plot <- ggplot(telo_data, aes(Population, ScaledTelomericReads, fill = Population)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", color = "red", size = 2) +
  scale_fill_manual(values = my_palette) +
  guides(fill = "none") +
  labs(x = "Populations", y = "Scaled Telomeric Reads") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  my_theme

ggsave("population_boxplot.png", allpop_plot, width = 12, height = 8, dpi = 300)
allpop_plot
```

---

### Gender Comparison

```{r}
gender_plot <- ggplot(telo_data, aes(Gender, ScaledTelomericReads, fill = Gender)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", color = "red", size = 3) +
  scale_fill_manual(values = c("female" = "lightseagreen", "male" = "lightpink2")) +
  guides(fill = "none") +
  labs(x = "Gender", y = "Scaled Telomeric Reads") +
  my_theme

ggsave("gender_boxplot.png", gender_plot, width = 8, height = 6, dpi = 300)
gender_plot
```

---

### Population × Gender 

```{r}
pop_gender_plot <- ggplot(telo_data, aes(interaction(Population, Gender), ScaledTelomericReads, fill = Gender)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", color = "red", size = 2) +
  labs(x = "Population × Gender", y = "Scaled Telomeric Reads") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  my_theme

ggsave("population_gender_boxplot.png", pop_gender_plot, width = 14, height = 8, dpi = 300)

pop_gender_plot
```

---

## Statistical Analysis

# The following tests were done to get an overview of the distribution and variability of telomere content

```{r}
summary(telo_data$ScaledTelomericReads)

telo_data %>%
  summarise(
    mean = mean(ScaledTelomericReads),
    median = median(ScaledTelomericReads),
    sd = sd(ScaledTelomericReads),
    min = min(ScaledTelomericReads),
    max = max(ScaledTelomericReads)
  )
```
---

## Normality Checks

```{r}
# Shapiro-Wilk test for all samples across all populations
# H0: A sample comes from a normal distribution
# H1: A sample DOES NOT come from a normal distribution

shapiro.test(telo_data$LogTelomere)

#Since Shapiro-Wilk is suitable for smaller sample size and we are working with larger sample size, oped for one-sample Kolmogorov-Smirnov test

#H0: The sample comes from the specified (e.g. normal) distribution. 
#H1: The sample DOES NOT come from a normal distribution

ks.test(telo_data$LogTelomere, "pnorm",
        mean = mean(telo_data$LogTelomere),
        sd = sd(telo_data$LogTelomere))
```

---
## QQ Plot

```{r}

qqnorm(telo_data$LogTelomere)
qqline(telo_data$LogTelomere, col = "blue")

# The points are generally close to the line in the middle portion, which suggests that the central values of the data roughly follow a normal distribution. However, there is some deviation at both ends (tails) of the plot, where the points diverge from the line. This suggests that the data have heavier tails than a normal distribution, implying more extreme values than expected under normality.

# Since the data is reasonably normally distributed, we used the parametric tests for our analyses.

```
---

## Homogeneity of Variance (Population)

```{r}
# Checking for Homogeneity of variance before doing an ANOVA, since it's an assumption with one -way ANOVA.

# H0: The population variances are equal
# H1: The population variances are not equal
# Reject the null hypothesis: If the p-value is less than a significance level of 0.05

leveneTest(LogTelomere ~ Population, data = telo_data)

#Since the p-value is much less than 0.05, Levene’s test suggests that the assumption of homogeneity of variance is violated. This means the variances across the groups (populations) are significantly different.
```

---

## Aim 1: Population Differences
# Que: Are there significant differences in telomeric content across diverse human populations?

# Since the homogeneity of variances assumption is violated, we considered using a one-way WELCH ANOVA test when doing multiple group comparisons
### Welch ANOVA

```{r}
#H0: There is no difference between the means of the populations
#H1: There is at least one population mean that is different from the others. 

oneway.test(LogTelomere ~ Population, data = telo_data)
```

### Post-hoc (Games-Howell)
# Perform Games- Howell test to identify which specific populations differ from each other.

```{r}
# H0: There is no difference in the mean values between the two groups.
# H1: There is a difference in the mean values between the two groups. 

games_howell_test(telo_data, LogTelomere ~ Population)
```

---

## Aim 2: Superpopulation Differences
# Que: Do the telomere content vary among different superpopulations?

```{r}

# Checking for Homogeneity of variance before doing an ANOVA, since it's an assumption with one -way ANOVA.

# H0: The superpopulation variances are equal
# H1: The superpopulation variances are not equal
# Reject the null hypothesis: If the p-value is less than a significance level of 0.05

leveneTest(LogTelomere ~ Superpopulation, data = telo_data)

#Since the p-value is much less than 0.05, Levene’s test suggests that the assumption of homogeneity of variance is violated. This means the variances across the superpopulations are significantly different.

#Since the homogeneity of variances assumption is violated, we considered using a one-way WELCH ANOVA test when doing multiple group comparisons

#H0: There is no difference between the means of the superpopulations
#H1: There is at least one superpopulation mean that is different from the others. 

oneway.test(LogTelomere ~ Superpopulation, data = telo_data)

games_howell_test(telo_data, LogTelomere ~ Superpopulation)
```

---

## Aim 3 & 4: Gender Differences

# Que: Do the telomere content vary between self-reported sexes within and across populations?

### Overall Gender

```{r}
# Checking for Homogeneity of variance before doing an ANOVA, since it's an assumption with one -way ANOVA.

# H0: The gender variances are equal
# H1: The gender variances are not equal
# Reject the null hypothesis: If the p-value is less than a significance level of 0.05

leveneTest(LogTelomere ~ Gender, data = telo_data)

# H0: There is no difference in the means of the two groups.
# H1: There is a difference in the means of the two groups.

t.test(LogTelomere ~ Gender, data = telo_data)
```

---

### Gender Differences Within Each Population 

```{r}
gender_results <- telo_data %>%
  group_by(Population) %>%
  summarise(p_value = t.test(LogTelomere ~ Gender)$p.value)

gender_results %>%
  mutate(
    bonferroni = p.adjust(p_value, method = "bonferroni"),
    fdr = p.adjust(p_value, method = "fdr")
  )

gender_results
```

---

## Conclusion
# This analysis demonstrates significant variation in telomere content across populations and superpopulations, while showing no strong global sex-based differences. Population-specific patterns may still exist and warrant further investigation.
